\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-30 paper.tex
Edits from Yan

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
ARGs are classically defined by the graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of sample,
with the differential inheritance structure encoded by the
crossover breakpoints associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of modern ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show this can fully capture the richness of modern large-scale datasets
and is compatible with modern simulation and inference methods.
Moreover, we show that, unlike the eARG encoding which requires
a precise estimate of all recombination events,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
text for details of these methods.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines indicate breakpoints between
local trees.
}
\end{figure}

@@

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume a
mutation rate of $5.49\times10^{-9}$
and recombination rate of $8.4\times10^{-9}$ per site per generation
% NOTE: we should be citing the original sources here too.
\citep{adrion2020community,lauterbur2023expanding},
% Note: would have been simpler to use the stdposim estimate
% of 1.7M from Li et al 2006. (not that it matters)
and constant effective population size of $10^6$.
Full details of the software used and code to generate these
figures are available on GitHub (see Data Availability).

% Here we describe recombination features
Considering~Fig.~\ref{fig-inferred-args}, we can see that
there is considerable variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the 7 required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly 2 parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of 2-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
% Several of these nodes have more than two
% % parents (in red), representing multiple recombinations inferred to have
% occurred in an unresolvable order.
% Nevertheless, the number of breakpoints in
% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
% that several multiple-parent nodes must share breakpoint positions.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.
% The sharing
% of breakpoint locations between these nodes indicates that in \tsinfer\ and
% \relate\ ARGs it may be difficult to condense each transition between trees
% into a set of simple SPR moves.
% Overall, Fig.~\ref{fig-inferred-args} shows
% that \tsinfer\ and \relate\ ARGs contain a level of detail about recombination
% that lies somewhere between a sequence of unrelated local trees on one extreme
% and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
Shading within nodes in Fig.~\ref{fig-inferred-args} indicates the
fraction of the node's span over which it is coalescent
(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
in Fig.~\ref{fig-inferred-args}A,B are clear because there is
no local coalescence at these recombination nodes
(they are ``ARG unary'', and so local coalescence is impossible).
The significant number of partially shaded nodes in
Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
\kwarg, \argweaver\ and \tsinfer\ all contain locally unary nodes.
Another difference between methods highlighted in this figure
is the presence of polytomies, which only
\tsinfer\ creates. The most obvious example involves
involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
which happen happen to have identical sequences.
Because \kwarg, \argweaver, and \relate\ require
bifurcating trees by design, each picks an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A,B disagrees
with Fig.~\ref{fig-inferred-args}D in this order).
% JK: I don't quite get this, so readers probably won't either. Let's
% stick a few clear messages.
% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
% that have more than two children hints at considerable uncertainty in
% the order of branching elsewhere in the ARG.

@@

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess,
it is not clear that generating many guesses
when sample sizes are large will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of likely ARGs.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference, and to
incorporate uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work. Fully utilising such uncertainty in downstream
applications is also vital, of course.

@@

In eukaryotic organisms, ARGs are a result of the cellular processes of mitosis
(which leads to common ancestry and coalescence)
and meiosis (which leads to recombination, including both crossovers and gene conversion).
Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
hermaphrodite eukaryote with partially overlapping generations. 
% It may be worth putting in something here about events being a slippery concept
% (we have already mentioned this at the end of Section 3
Here, an event is not represented by a specific
genome: rather, a genome can be associated with (or ``tag'') an event above or below it.
The gARG representation allows any such genome to be arbitrarily distant in time or number of cell
divisions from the event itself (see Section~\ref{sec-precision}).

Thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes, and these the lines pass through gARG nodes (green chromosomal regions).
A number of these nodes have only one parent and one child in the ARG; these would often be omitted
(taken as implicit) in a minimally simplified gARG.

@@

It is worth noting that ``true'' polytomies in ARGs are a feature of model assumptions
rather than biological reality. Indeed, given a large enough mutation rate
(or single cell sequencing at a sufficiently detailed level) it should be possible to
distinguish sequential rounds of mitosis within a single multicellular organism.
For example, a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
the right hand gamete. For an ARG that has been accurately inferred from real data,
a polytomy is therefore making a (deliberately) imprecise statement about the order
of coalescence, rather than reflecting of a single instantaneous event; this
interpretation fits naturally into the gARG framework.

>>>>

\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG).
% Be specific about what we're talking about.
ARGs are classically defined by the graph of
% We can play a bit loose here, please don't change to "common ancestor"
coalescence and recombination events in the history of a sample,
with the differential inheritance structure encoded by the
crossover breakpoints associated with recombination events.
We show that this ``event ARG'' (eARG) encoding
is limited in the patterns of genetic inheritance that can be represented,
being insufficient to describe the outputs of modern ARG simulation
and inference methods,
and ill-suited to capturing the complex patterns of
relatedness that exist in contemporary datasets.
We propose an alternative ``genome ARG'' (gARG) encoding
expressed in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance.
We show that the gARG encoding can fully capture the richness of modern large-scale datasets
and is compatible with modern simulation and inference methods.
Moreover, we show that, unlike the eARG encoding which requires
% TODO rephrase "to be detailed" is ambiguous
all recombination events to be detailed,
the gARG encoding enables fine-grained
levels of precision about recombination to be represented,
opening up a rich set of additional details.
\end{abstract}

@@

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} DNA sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods:
(A) \kwarg; (B) \argweaver; (C) \tsinfer; and (D) \relate. See the
text for details of these methods.
Edge colours indicate time of the edge's child node
(lighter: older; darker: younger).
Vertical and horizontal positions of graph nodes are arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
Bottom row graphics show the genome positions, relative to the start of
the ADH gene, for each graph edge from the corresponding ARG. Edge intervals
are drawn as horizontal lines, stacked in time order (edges with youngest
children at the bottom); vertical dashed lines denote breakpoints between
local trees.
}
\end{figure}

@@

Inferred ARGs are based on the \citet{kreitman1983nucleotide} dataset,
a standard benchmark in the classical ARG
literature.
It consists of 43 biallelic SNPs spanning 2.4Kb of the \emph{D.~melanogaster} ADH
locus on chromosome 2L. Where required for inference purposes we assume
mutation and recombination rates of $5.49\times10^{-9}$
and $2.40463\times10^{-9}$ per site per generation
% NOTE: we should be citing the original sources here too.
\citep{schrider2013rates,comeron2012many}
and a constant effective population size of $1720600$ \citep{li2006inferring}, all
values being taken from the \texttt{stdpopsim} library
\citep{adrion2020community,lauterbur2023expanding}.
Full details of the software used and code to generate these
figures are available on GitHub (see Data Availability).

% Here we describe recombination features
Considering Fig.~\ref{fig-inferred-args}, we can see that
there is substantial variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the 7 required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly 2 parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of 2-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and number of breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
% Several of these nodes have more than two
% % parents (in red), representing multiple recombinations inferred to have
% occurred in an unresolvable order.
% Nevertheless, the number of breakpoints in
% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
% that several multiple-parent nodes must share breakpoint positions.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.
% The sharing
% of breakpoint locations between these nodes indicates that in \tsinfer\ and
% \relate\ ARGs it may be difficult to condense each transition between trees
% into a set of simple SPR moves.
% Overall, Fig.~\ref{fig-inferred-args} shows
% that \tsinfer\ and \relate\ ARGs contain a level of detail about recombination
% that lies somewhere between a sequence of unrelated local trees on one extreme
% and an ARG with precisely specified recombination events on the other.

% Here we move on to coalescence
Shading within nodes in Fig.~\ref{fig-inferred-args} indicates the
fraction of the node's span over which it is coalescent
(Section~\ref{sec-locally-unary-edges}). For example, brown nodes
in Fig.~\ref{fig-inferred-args}A,B are clear because there is
no local coalescence at these recombination nodes
(they are ``ARG unary'', and so local coalescence is impossible).
The significant number of partially shaded nodes in
Fig.~\ref{fig-inferred-args}A,B and C demonstrates that
\kwarg, \argweaver\ and \tsinfer\ all contain locally unary nodes.
Another difference between methods highlighted in this figure
is the presence of polytomies, which only
\tsinfer\ creates. The most obvious example
involves nodes \noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F},
which happen to have identical sequences.
Because \kwarg, \argweaver, and \relate\ require
bifurcating trees by design, each picks an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args}A and B disagree
in this order, and Fig.~\ref{fig-inferred-args}D even shows different
orders in different trees).
% JK: I don't quite get the commented out sentence below, so readers probably
% won't either. Let's stick to a few clear messages.
% More generally, the number of nodes in Fig.~\ref{fig-inferred-args}C
% that have more than two children hints at considerable uncertainty in
% the order of branching elsewhere in the ARG.

@@

Recent large-scale methods have simplified the inference problem by
making a single, deterministic best-guess at ARG
inference~\citep{kelleher2019inferring,speidel2019method,zhan2023towards}.
Even under strict parsimony conditions and for small sample sizes, the
% Citation? This is something people have figured out, right?
number of plausible ARGs compatible with a given dataset is vast.
Thus, although it is clearly an oversimplification to arbitrarily
choose \emph{one} best guess,
it is not clear that generating many guesses
when sample sizes are large will achieve much.
At the scale of millions of samples,
we could only ever explore the tiniest corner of the incomprehensibly
large space of likely ARGs.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference, and to
incorporate uncertainty encountered during inference into the returned ARG.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work. 
% Obvious, but nobody is doing it, so needs to be said. 
% TODO rephrase to say this is an open question.
Fully utilising such uncertainty in downstream
applications is also vital.

@@

In eukaryotes, ARGs are a result of the cellular processes of mitosis
(which leads to common ancestry and coalescence)
and meiosis (which leads to recombination, including both crossovers and gene conversion).
Fig.~\ref{fig-simplification} shows a schematic of the events (coloured red and blue squares) and
the genomes (chromosome icons) that occur in the cellular germline of a simplified, diploid multicellular
hermaphrodite eukaryote with partially overlapping generations. 
% It may be worth putting in something here about events being a slippery concept
% (we have already mentioned this at the end of Section 3
Here, an event is not represented by a specific
genome. Rather, a genome can be associated with (or ``tag'') an event above or below it (and as argued in
Section~\ref{sec-precision}, the gARG representation allows these associated genomes to be arbitrarily distant
in time or number of cell
divisions from the event itself).

Thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes, and these lines pass through gARG nodes (green chromosomal regions).
A number of these nodes have only one parent and one child in the ARG; these would often be omitted
(taken as implicit) in a minimally simplified gARG.

@@

Note that mutations can occur along any lineage in Fig.~\ref{fig-cell-lines}. This means, for example, 
that a mutation in the first cell division of $D_{10}$ could be shared between
the two gametes produced by the cells in the left half of $D_{10}$ but not shared by
the right hand gamete. With a high enough germline mutation rate (or equivalently, a long enough genome), each
round of mitotic genome duplication within a single multicellular organism could
potentially be distinguished.
This implies that in ARGs that have been inferred from real data, polytomies represent a lack
of mutational information (rather than marking a single ``multifurcation event'').
The ability to represent polytomies, and conceptualise them as a (deliberately) imprecise statement about the order
of coalescence within and between individuals,
fits naturally into the gARG framework presented in this paper.
